Distributed Topology Discovery in Self-Assembled Nano Networks

### 2014-02-14 trunk/PAPER/disr.tex
monteleone's fixes
DiSR sequence pics and dbs
Misc fixes

%------------------------------------------------------------------------------
\section{Preliminary concepts}
%------------------------------------------------------------------------------
\label{sec:disr_concepts}

\subsection{Basic Idea}
The main concept behind the kind of turn prohibition we want to
achieve is the \emph{Segment}: it is basically a path of consecutive
nodes and links. A segment $S_1$ starts with a link attached to a node which
belongs to different segment $S_2$, ending with a link attached to a
node belonging to a different segment $S_3$. In other words, a segment
a path between connecting two other different segments.A exception is
the first segment established in the network, called \emph{starting
segment} which is  a loop beginning/ending on a particular node
defined as \emph{bootstrap node}.
The idea is to partition the network in a set of disjoint segments, and then
placing a turn restriction within each segment. It has been proved~\cite{mejia_ipdps06}
that such a set of turn prohibitions guarantees deadlock freedom and
while preserving connectivity of the network.

Describing the execution phases of DiSR from a top level point-of-view
is basically similar to the classic SR approach. This can be useful to
give an initial idea of the approach, however, some essential
differences should be pointed out. First of all no centralized entity is
globally aware of what is going on: the status of the DiSR execution
is collectively distributed among the nodes; second, no defect map and/or
topology graph is available as input, thus, the topology has to be
discovered \emph{while} segments are created; finally, at the end of the
execution, no segment list is created: Each node is only aware of
belonging to some segment, ignoring the presence of other nodes
in the same segment and even the presence of other segments in the
network.

Roughly speaking, the execution of can be described as the execution
of the following phases: (1) Injection of the DiSR process from upper layer to set a bootstrap
node (2) Bootstrap node broadcasting to create the first segment of the subnet
(3) Parallel requests starting from assigned node to discover other
segments. The following subsection describe the data structure and
execution model required at each node. (4) Turn proibition in each
confirmed segment.
%\item Draining of remaining packet due timeouts

%Once again, it should be pointed out that nowhere in the network there
%is such a global vision of these exection phases. A more detailed
%description of the exection model at each node is discussed in the
%next section.

\subsection{Signal Types Required}

The DiSR approach works with a distributed mechanism which is build
upon an exchange of small packets containing the following fields:
\begin{itemize}
\item{packet\_type}: encodes the meaning of the control message respect
to the whole DiSR process.
\item{seg\_ID}: the id of the segment associated to the DiSR control message
\item{src\_id}: the id of the node that initially originated the
packet
\end{itemize}

As regards packet\_type, we can have the following control packets:
\begin{itemize}
\item{\texttt{STARTING\_SEGMENT\_REQUEST}}: Injected by the bootstrap
node to when searching for the first segment. 
\item{\texttt{STARTING\_SEGMENT\_CONFIRM}}: used when establishing
the starting  segment. 
\item{\texttt{SEGMENT\_REQUEST}}: used to search candidates for a segment (not the
first)
\item{\texttt{SEGMENT\_CONFIRM}}: used to establish a segment. 
\item{\texttt{SEGMENT\_CANCEL}}: used to cancel the process of searching a segment along a
specific link.
\end{itemize}

A quantitative analysis of the resources needed to implement these
fields is presented in Section~\ref{sec:implementation} when
discussing the impact of the DiSR control logic and storage on
hardware implementation of the node.

\subsection{DiSR Data Structures}
\label{ssec:disr_dstruct}

@@

The \emph{LED} is like a snapshot of the DiSR algorithm at each node,
consisting in the following variables:
\begin{itemize}
\item{\emph{segID}}:a value used to specify the segment to which the
node has been assigned or is candidate for being assigned.
\item{\emph{visited}}: a boolean value. When \emph{true}
a \emph{segID} different from \texttt{NULL} specifies the segment 
to which the node has been assigned. 
% TODO: the following is true, but complicates things at this stage
%Note that a value different from \emph{NULL} does not
%necessarily mean the node has been assigned to some segment (e.g. when
%searching starting segment is set \emph{visited} by default even if not yet
%assigned to any segment). 
\item{\emph{tvisited}}: a boolean value. If \emph{true}, the node is
being considered as candidate for a segment, and the \emph{segID} value
specifies the segment id for which the node is candidate. 
\item{\emph{link\_visited[]}}: an array
of values representing information about attached links, that is, the
segID of the segment owning each link. When \texttt{NULL}, the corresponding link has not yet been
assigned.
%TDB: do we need to introduce some value for the
%“bridge” status ? (e.g. a node connecting the terminal node of a
%subnet with the starting one of another) 
\item{\emph{link\_tvisited[]}}: an array of
values representing information about attached links, that is, the \emph{segID} of
the segment for which the link is candidate. When \texttt{NULL}, the link is
currently not \emph{tvisited}.  
\end{itemize}
%\item{\emph{starting}}: boolean, \emph{true} if the node is a
%starting node from which the whole process was initiated.
%\end{itemize}

@@

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the DiSR algorithm.
The node is not marked as \emph{visited}/\emph{tvisited}.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
% TODO: this and all the other details on the items below should be
% checked and moved elsewhere for a future long paper
%This is required because a
%\texttt{STARTING\_SEGMENT\_REQUEST} packet should be injected in order to start
%the whole process.
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
%This happens when all
%the following conditions are matched: the node is marked as \emph{visited},
%that is \emph{visited} = TRUE 
%TODO: \emph{segID} is set to some value reflecting
%the segment request sent there is a link that is being investigated by
%the node, that is a tvisited\_link marked with the same id of the node 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
with id segid, not being itself the node from which the find process
was started.

%The following conditions should be matched: \emph{tvisited} =
%true, with segID=X different from node's id.  Two links set as
%\emph{link\_tvisited} with id X. Let’s refer them as  links 'i' and 'j'. Then
%link\_tvisited[i] was set = X when the find process from an adjacent
%node reached the current node, while link\_tvisited[j] has been set to
%X when the node itself started to investigate its free links for the
%segment request with id X. Note that while investigating its attached
%links, if the find process fails along path 'j', the current node can
%investigate another free node 'k' if suitable.  
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment. 
%Two main differences: Can confirm
%\texttt{STARTING\_SEGMENT\_CONFIRM} with the same \emph{segID} when receiving a normal
%\texttt{SEGMENT\_REQUEST}: Overwrites its status setting it to \emph{Candidate} for the
%segID frees the links previously set as \emph{tvisited} during flooding
\item{\emph{Assigned}}: a node for which the segment has been determined.  The
segment \emph{segID} attribute value is set to some id X different from \texttt{NULL}.
%The \emph{visited} value is \emph{true} NOTE: an \emph{Assigned} state is a quick temporary
%state, since it becomes \emph{ActiveSearching} if a not (\emph{visited}/tvisited)
%link is suitable for searching new segments (see next\_visited
%procedure on paper) 
\end{itemize}

\subsection{DiSR Execution Model}

In Figure~\ref{fig:state_machine} are depicted the main events
representing the execution model at each node. The DiSR control
packets described in the previous subsection trigger node events,
which eventually change $DBS$ status according to the current value of
$LED$ variables and the type of the packet received. The main DiSR
phases, together with a reference to the particular status transition
shown in Figure~\ref{fig:state_machine} are described in the
following.

\begin{figure}
  \centering
    \includegraphics[width=0.40\textwidth]{pictures/state_machine.eps}
  \caption{DiSR node execution model}
  \label{fig:state_machine}
\end{figure}
\begin{itemize}

\item{\textbf{Injecting bootstrap request}}: all nodes have an initial
\emph{DBS} status \emph{Free}, except for a
 node with status \emph{Bootstrap}, set by some signal from an
upper layer via (a). When starting, this bootstrap node 
changes its status to \emph{ActiveSearching} (b), injecting a
\texttt{STARTING\_SEGMENT\_REQUEST} across one of its free links.

\item{\textbf{Flooding}}: Each node receiving a \texttt{STARTING\_SEGMENT\_REQUEST},
when $Free$ forwards it to its free links using a flooding mechanism and
becoming $CandidateStarting$ (g).
Each of its free links is then marked as \emph{tvisited} with the segment id
associated to the request. Note that a node that has already received a
\texttt{STARTING\_SEGMENT\_REQUEST} packet can simply ignore further packets
associated to the same request, having already contributed to the
flooding. 
%Note also that each of these packets has a max\_segment\_hops
%field to prevent packets from travelling undefinitively.

\item{\textbf{Confirming the Starting Segment}}: when the \texttt{STARTING\_SEGMENT\_REQUEST} packet reaches
the bootstrap node (from a different link, of course), the starting
segment is found. Then the bootstrap node sends back a \texttt{STARTING\_SEGMENT\_CONFIRM}
packet along the link from which it received the
\texttt{STARTING\_SEGMENT\_REQUEST} and becomes $Assigned$ (c). Each
node receiving the confirm do same by changing its own status
to \emph{Assigned} (l). So the confirmation packet is sent back from
node to node and the starting segment is created.

Note: A node being \emph{CandidateStarting}, when receiving a simple
(not starting) \texttt{SEGMENT\_REQUEST} , can simply cancel its
previous \emph{CandidateStarting} status and set itself as
$Candidate$ for that request (h), since this new
request means that the starting segment has already been found.

@@

\item{\textbf{Injecting other requests}}: Each node in the
\emph{Assigned} status can initiate a search for a segment, by sending
a \texttt{SEGMENT\_REQUEST} across one of its free links (d). Note
that since this is not the starting segment in this case the packet
should not reach the initiator, but just another \emph{Assigned} node.

\item{\textbf{Setup of a segment}}: The find process is successful
when an already \emph{Assigned} node receives the
\texttt{SEGMENT\_REQUEST} packet. Then, a \texttt{SEGMENT\_CONFIRM} is
sent back along the path that originated the request (e), while the
node remain $Assigned$ (since it could confirm more segments). Each node
previously set as $Candidate$ for that segment id, when receiving the
confirm packet changes its status to $Assigned$ (m) and forwards back
the same confirm until it reaches the initiator of the request, which
changes from $ActiveSearching$ to $Assigned$ (c).

\item{\textbf{Failing while searching a segment}}: a node received a \texttt{SEGMENT\_REQUEST} packet but
matched one of two the following conditions: the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})(f); the node
is candidate for another find process(i).  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the proper link. 
%Each of the subsequent nodes will forward this \texttt{SEGMENT\_CANCEL}
%only if no more free links can be explored starting from them.
\end{itemize}

\subsection{Intra-node vs Inter-node parallelism}
In the processes described above, we assumed that each node in the
READY\_SEARCHING state can start a find segment process by injecting a
\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when defining DiSR
should be made in order to decide whether or not the node must
actually perform this action. 
The point above is is strictly related to the general question: which
level of “parallelism” should we allow in DiSR ? The adopted approach
is that, although the nature of DiSR itself is intrinsicly parallel,
the use of parallelism makes things work in a more complex way. In
other words, DiSR is parallel when needed, but it does not exploit
parallelism as an “improving feature”. When not needed, things should
be serialized. For example: when a node is “ready searching” could
start several find segment process associated to the same segment, one
for each free link. But serializing this by investigating the free
links in order could be a simplest solution. We refer to this saying
that we avoid intra-node parallelism.  Note that while intra-node
parallelism can be avoided, trying to avoid inter-node parallelism
could be more complex than allowing the parallelism itself. Imagine
for example the effort of trying to coordinate nodes so that an unique
finding node process is actually running in the whole subnet. Thus, in
contrast to the intra-node parallelism, the inter-node parallelism is
a structural property of the DiSR algorithm and should not be avoided.
%------------------------------------------------------------------------------

>>>>

%------------------------------------------------------------------------------
\section{\disr{} Overview}
%------------------------------------------------------------------------------

\label{sec:disr_concepts}
The main concept behind the kind of turn prohibition we want to
achieve is the \emph{segment}. A segment $S_y$ is basically a path of consecutive
nodes and links, starting with a link attached to a node belonging to
a different segment $S_x$ and ending with a link attached to a
node of another segment $S_z$. In other words, a segment
a path between connecting two other different segments. The example
in Figure~\ref{fig:segments} shows the segment with id $1.4$, starting
with a link connected to the node $9$, the going trought nodes
$4$,$3$,$2$ and ending with the link connected to the node $1$. Other
segments depicted are $9.2$ (link,node $10$,link) and $9.3$
(link from $9$ ,$8$,link,$7$,link,$6$,link,$5$,link to the $1$). Note
also that start/ending link can coincide and a segment can also be
formed by one single link (e.g. in $7.2$ or $5.4$).
An exception to the rule that a segment must connect two other
segments is the first segment established in the network, called
\emph{starting segment} which is a loop beginning/ending on the same
node.

\begin{figure}
\centering
    \includegraphics[width=0.25\textwidth]{pictures/network_slice.eps}
  \caption{Example of Segments}
  \label{fig:segments}
\end{figure}

A detailed description of the \disr{} execution model as depicted in
the Figure~\ref{fig:disr_events} will be discussed in
Subsection~\ref{sub:phases}. Roughly speaking, the execution of can be described as the execution
of the following phases: 
\begin{enumerate}
\item Injection of the \disr{} process from an upper layer to set a \emph{bootstrap
node} 
\item Broadcasting and confirmation to create the first segment of the subnet
\item Parallel requests starting from assigned node to discover other
segments. 
\item Turn proibition in each confirmed segment.
\item Draining of remaining packet due timeouts
\end{enumerate}

\section{Detailed Description of \disr{}}
\subsection{Message Types Required}

The \disr{} approach works with a distributed mechanism which is build
upon an exchange of small packets containing the following fields:
\begin{itemize}
\item \emph{packet\_type}: encodes the meaning of the control message
\item \emph{seg\_ID}: the id of the segment associated to the \disr{} control message
\item \emph{src\_id}: the id of the node that originated the packet
\end{itemize}

As regards \emph{packet\_type} field, we can have the following control packets:
\begin{itemize}
\item{\texttt{STARTING\_SEGMENT\_REQUEST}}: Injected by the bootstrap
node to when searching for the first segment. 
\item{\texttt{STARTING\_SEGMENT\_CONFIRM}}: used when establishing
the starting  segment. 
\item{\texttt{SEGMENT\_REQUEST}}: used to search candidates for a segment (not the
first)
\item{\texttt{SEGMENT\_CONFIRM}}: used to establish a segment. 
\item{\texttt{SEGMENT\_CANCEL}}: used to cancel the process of searching a segment along a
specific link.
\end{itemize}

\subsection{\disr{} Data Structures}
\label{ssec:disr_dstruct}

@@

The \emph{LED} is like a snapshot of the \disr{} algorithm at each node,
consisting in the following variables:
\begin{itemize}
\item{\emph{segID}}:a value used to specify the segment to which the
node has been assigned or is candidate for being assigned.
\item{\emph{visited}}: a boolean value. When \emph{true}
a \emph{segID} different from \texttt{NULL} specifies the segment 
to which the node has been assigned. 
% TODO: the following is true, but complicates things at this stage
%Note that a value different from \emph{NULL} does not
%necessarily mean the node has been assigned to some segment (e.g. when
%searching starting segment is set \emph{visited} by default even if not yet
%assigned to any segment). 
\item{\emph{tvisited}}: a boolean value. If \emph{true}, the node is
being considered as candidate for a segment, and the \emph{segID} value
specifies the segment id for which the node is candidate. 
\item{\emph{link\_visited[]}}: an array
of values representing information about attached links, that is, the
segID of the segment owning each link. When \texttt{NULL}, the corresponding link has not yet been
assigned.
%TDB: do we need to introduce some value for the
%“bridge” status ? (e.g. a node connecting the terminal node of a
%subnet with the starting one of another) 
\item{\emph{link\_tvisited[]}}: an array of
values representing information about attached links, that is, the \emph{segID} of
the segment for which the link is candidate. When \texttt{NULL}, the link is
currently not \emph{tvisited}.  
\end{itemize}
%\item{\emph{starting}}: boolean, \emph{true} if the node is a
%starting node from which the whole process was initiated.
%\end{itemize}

@@

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the \disr{} algorithm.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
%This happens when all
%the following conditions are matched: the node is marked as \emph{visited},
%that is \emph{visited} = TRUE 
%TODO: \emph{segID} is set to some value reflecting
%the segment request sent there is a link that is being investigated by
%the node, that is a tvisited\_link marked with the same id of the node 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
, but not being itself the node from which the find process was started.

%The following conditions should be matched: \emph{tvisited} =
%true, with segID=X different from node's id.  Two links set as
%\emph{link\_tvisited} with id X. Let’s refer them as  links 'i' and 'j'. Then
%link\_tvisited[i] was set = X when the find process from an adjacent
%node reached the current node, while link\_tvisited[j] has been set to
%X when the node itself started to investigate its free links for the
%segment request with id X. Note that while investigating its attached
%links, if the find process fails along path 'j', the current node can
%investigate another free node 'k' if suitable.  
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment. 
%Two main differences: Can confirm
%\texttt{STARTING\_SEGMENT\_CONFIRM} with the same \emph{segID} when receiving a normal
%\texttt{SEGMENT\_REQUEST}: Overwrites its status setting it to \emph{Candidate} for the
%segID frees the links previously set as \emph{tvisited} during flooding
\item{\emph{Assigned}}: a node for which the segment has been determined.  
%The \emph{visited} value is \emph{true} NOTE: an \emph{Assigned} state is a quick temporary
%state, since it becomes \emph{ActiveSearching} if a not (\emph{visited}/tvisited)
%link is suitable for searching new segments (see next\_visited
%procedure on paper) 
\end{itemize}

A quantitative analysis of the resources needed to implement these
fields is presented in Section~\ref{sec:implementation} when
discussing the impact of the \disr{} control logic and storage on
hardware implementation of the node.

\subsection{Execution Model}
\label{sub:phases}

The \disr{} control packets described in the previous subsection trigger
node events, which eventually change $DBS$ status according to the
current value of $LED$ variables and the type of the packet received.
The main \disr{} phases, together with a reference to the particular
status transition involved are described in the following. As a
convention, we used single letters to refer to one of the \emph{DBS} transitions in
Figure~\ref{fig:state_machine} while the phases are visually
represented in~\ref{fig:disr_events}.

\begin{figure}
  \centering
    \includegraphics[width=0.25\textwidth]{pictures/dbs_updated.eps}
  \caption{\disr{} node execution model}
  \label{fig:state_machine}
\end{figure}

\begin{figure*}
\centering
\begin{tabular}{cccc}
\includegraphics[width=0.23\textwidth]{pictures/seq01.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq02.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq03.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq04.eps} \\
(a) & (b) & (c) & (d) \\
\includegraphics[width=0.23\textwidth]{pictures/seq05.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq06.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq07.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq08.eps} \\
(e) & (f) & (g) & (h) \\
\end{tabular}
\caption{Example of \disr{} segment setup: Nodes status are labeled as CandidateStarting($Cs$), Candidate ($C$), Free ($F$), Assigned ($A$) and Boostrap ($B$)}
\label{fig:disr_events}
\end{figure*}

\begin{itemize}
\item{\textbf{Injecting bootstrap request}}: all nodes have an initial
\emph{DBS} status set to \emph{Free}, except for a
 node with status \emph{Bootstrap}, set by some control signal from an
upper layer via $(a)$. When starting, this bootstrap node 
changes its status to \emph{CandidateStarting} $(b)$, injecting a
\texttt{STARTING\_SEGMENT\_REQUEST} across one of its free links
(Figure~\ref{fig:disr_events}-a).

\item{\textbf{Bootnode Broadcasting}}: Each node receiving a \texttt{STARTING\_SEGMENT\_REQUEST},
when $Free$, forwards it to its free links using a flooding mechanism and
becoming $CandidateStarting$ $(g)$.
Each of its free links is then marked as \emph{tvisited} with the segment id
associated to the request. Note that a node that has already received a
\texttt{STARTING\_SEGMENT\_REQUEST} packet can simply ignore further packets
associated to the same request, having already contributed to the
flooding (Figure~\ref{fig:disr_events}-b).
%Note also that each of these packets has a max\_segment\_hops
%field to prevent packets from travelling undefinitively.

\item{\textbf{Confirming the Starting Segment}}: when a \texttt{STARTING\_SEGMENT\_REQUEST} packet reaches
the bootstrap node (from a different link, of course), the starting
segment is found. Then the bootstrap node sends back a \texttt{STARTING\_SEGMENT\_CONFIRM}
packet along the link from which it received the
\texttt{STARTING\_SEGMENT\_REQUEST} and becomes $Assigned$ $(l)$. Each
node receiving the confirm do same by changing its own status
to \emph{Assigned}(Figure~\ref{fig:disr_events}-c). So the confirmation packet is sent back from
node to node and the starting segment is created (Figure~\ref{fig:disr_events}-d).

@@

\item{\textbf{Injecting Segment Requests}}: Each node in the
\emph{Assigned} status can initiate a search for a segment, by sending
a \texttt{SEGMENT\_REQUEST} across one of its free links $(c)$ (see
Figure~\ref{fig:disr_events}-e). 
Note that each \emph{Assigned} note can inject the request just after
becoming assigned, i.e. it is not needed that the whole confirmation
process in completed (Figure~\ref{disr_events}-f).
Note also that nodes previosly set as \emph{CandidateStarting}, when receiving 
a \texttt{SEGMENT\_REQUEST} can simply cancel their
status and set as $Candidate$ for that request, since this new
request means that the first starting segment has already been found $(h)$. 
\item{\textbf{Segment Confirmation}}: The find process is successful
when an already \emph{Assigned} node receives the
\texttt{SEGMENT\_REQUEST} packet. Then, a \texttt{SEGMENT\_CONFIRM} is
sent back along the path that originated the request
(Figure~\ref{fig:disr_events}-g), while the
node remain $Assigned$ (since it could confirm more segments). Each node
previously set as $Candidate$ for that segment id, when receiving the
a confirmation packet changes its status to $Assigned$ $(m)$ and forwards back
the same confirmation until it reaches the initiator of the request,
which then changes from $ActiveSearching$ to $Assigned$ $(d)$.

\item{\textbf{Failing while searching a segment}}: this happen when a node received a \texttt{SEGMENT\_REQUEST} packet but
matches one of two the following conditions: first, the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})$(f)$; the node
is already \emph{Candidate} with another segment id, i.e. another find
process.  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the incoming link (see at
the bottom of Figure~\ref{fig:disr_events}).
%Each of the subsequent nodes will forward this \texttt{SEGMENT\_CANCEL}
%only if no more free links can be explored starting from them.
\end{itemize}
